Labeling in Stucplots

### 2005-08-02 trunk/vcd/inst/doc/labeling.Rnw
Vignettes enhanced. labeling_text.Rd renamed to labeling_border.Rd
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@422 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

One of the major enhancements in vcd is the labeling of strucplots which now
has become much more powerful and flexible. Like the panel, the
shading, the spacing, and the legend, labeling is now completely modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \code{strucplot()} which then draws the labels. The labeling is 
well-separated from the actual plotting that occurs in the panel functions;
it actually only relies on the viewport tree produced by them, and the
`dimnames' attribute of the visualized table. The drawing of the
labels happens after the panel function has drawn the actual plot.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
As an example, consider \code{labeling\_list()}:

@@

\noindent Note that the last two levels of the `Survived' variable do
overlap. This can be addressed in several ways. The `brute force'
method is to enable clipping for the `Survivcal' dimension (see Figure \ref{fig:clipping}):

@@

\noindent A more sensible solution is to abbreviate the `Survival'
levels (see Figure \ref{fig:abbreviating}):

@@

\noindent The \code{abbreviate} argument actually takes a vector of
integers indicating the number of characters the levels should be
abbreviated to, but logical values are equivalent to 1. Another
possibility is to rotate the levels (see Figure \ref{fig:rotating}):

@@

We proceed with a few more `cosmetic' changes. 
A first simple, but effectful modification is to position
all labels and variables left-aligned, which is
also achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):

@@

\noindent Next, we show how to put all levels to the
bottom and right margins, and all variable names to the top and left
margins (see Figure \ref{fig:margins}):

@@

\noindent Now, we will add boxes to the labels and additionally
enable clipping (see Figure \ref{fig:boxes}):

@@

\noindent This is pretty close to what the \code{labels\_cboxed()}
wrapper does. Another variant is to put the variable names into the
same line as the levels (see Figure \ref{fig:labbl}):

@@

\noindent This is basically what \code{labeling\_lboxed()} produces,
and a similar design is used by the \code{doubledecker()}.

@@

\noindent In the case of narrow cells, the labels can be abbreviated 
(see Figure \ref{fig:cell2}):

>>>>

One of the major enhancements in vcd is the labeling of strucplots which now
has become much more powerful and flexible. Like the panel, the
shading, the spacing, and the legend, labeling is now completely modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \code{strucplot()} which then draws the labels. The labeling is 
well-separated from the actual plotting that occurs in the panel functions;
it actually only relies on the viewport tree produced by them, and the
`dimnames' attribute of the visualized table. The drawing of the
labels happens after the actual plot has been drawn by the panel function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
As an example, consider \code{labeling\_list()}:

@@

\noindent Note that the last two levels of the `Survived' variable do
overlap. This issue can be addressed in several ways. The `brute force'
method is to enable clipping for the `Survivcal' dimension (see Figure \ref{fig:clipping}):

@@

\noindent The \code{clip} parameter is passed to the labeling function
via the \code{labeling\_args} argument which takes a list of
parameters. \code{clip} itself takes a vector of logicals (one for
each dimension). Almost all vectorized arguments in the strucplot
framework can be abbreviated in the following way: unnamed components
(or the defaults, if there are none) are recycled as needed, but
overridden by the named components. Here, the default is \code{FALSE},
and therefore clipping is enabled only for the \code{Survived} dimension.
A more sensible solution to the overlap problem is to abbreviate the `Survival'
levels (see Figure \ref{fig:abbreviating}):

@@

\noindent The \code{abbreviate} argument actually takes a vector of
integers indicating the number of characters the levels should be
abbreviated to, but logical values are equivalent to 1. Abbreviation
is performed using {\sf R}'s \code{abbreviate()} function. Another
possibility is to rotate the levels (see Figure \ref{fig:rotating}):

@@

We now proceed with a few more `cosmetic' features.
A first simple, but effectful modification is to position
all labels and variables left-aligned, which can
also be achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):

@@

\noindent Note that obviously we need to change the justification to \code{"left"} as well. Next, we show how to put all levels to the
bottom and right margins, and all variable names to the top and left
margins (see Figure \ref{fig:margins}):

@@

\noindent The \code{"tl\_foo"} (``top left'') arguments are \code{TRUE} by default. 
Now, we will add boxes to the labels and additionally
enable clipping (see Figure \ref{fig:boxes}):

@@

\noindent The values to \code{boxes} and \code{clip} are recycled for
all dimensions. The result is pretty close to what the \code{labels\_cboxed()}
wrapper does. Another variant is to put the variable names into the
same line as the levels (see Figure \ref{fig:labbl}):

@@

\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
labels'') is a vector of logicals indicating the side for the
variable names. The resulting layout is basically what \code{labeling\_lboxed()} produces,
and a similar design is used by the \code{doubledecker()}.

@@

\noindent In the case of narrow cells, it might be useful to
abbreviate labels and/or variable names and turn off clipping (see Figure \ref{fig:cell2}):
